---
title: "MonteCarloPresidency"
author: "Grant Corkren"
date: "2025-01-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This is a just a mini-project for finding the distribution of electoral college votes.

Simulated polling data will be based on the 2024 election results proportions and taking a random n=1 sample for each state from standard normal distribution. That will then be transformed into a simulated proportion. The simulated sample sizes will be calculated by dividing the total vote count of each state by ten thousand. 

Polling data where sample sizes of n > 30 for each state would be the appropiate methodology. In that case the CLT would apply to the mean of proportions for each state. That would give it a normal distribution for each state where a monte carlo simulation could be applied. That method will be applied, but it implies independence among states. A multivariate normal distribution where the covariance matrix is built on previous elections will also be used. 

```{r}
setwd("~/MY_R")
# read in data
library(openxlsx) 
library(ggplot2)
sim_data <- read.xlsx("2024results.xlsx")
head(sim_data)
```
```{r}
# total votes column divided by 10,000
sim_data <- sim_data %>%
  mutate(
    total_votes = Trump + Harris,          
    simulated_sample_size = round(total_votes / 10000)  
  )

```

```{r}
# simulated proportions
set.seed(1)  
sim_data <- sim_data %>%
  rowwise() %>%
  mutate(
    simulated_proportion = rnorm(
      n = 1,                                 
      mean = REPUBLICANPERCENTAGE,          
      sd = sqrt((REPUBLICANPERCENTAGE * (1 - REPUBLICANPERCENTAGE)) / simulated_sample_size) 
    )
  ) %>%
  ungroup()

head(sim_data)
```

```{r}
sim_data <- sim_data %>%
  mutate(
    variance = (simulated_proportion * (1 - simulated_proportion)) / simulated_sample_size
  )
head(sim_data)
```

```{r}
tail(sim_data)
```


Two approaches 
  1) Assume Independence  
  2) use past data to get covariance matrix


# Assumed Independence

```{r}
# monte carlo simulation
set.seed(1)  
n_simulations <- 10000  


simulation_results <- replicate(n_simulations, {
  simulated_values <- pmin(pmax(rnorm(nrow(sim_data), mean = sim_data$simulated_proportion, 
                                      sd = sqrt(sim_data$variance)), 0), 1)
  
  total_electoral_votes <- sum(sim_data$Elec_votes[simulated_values > 0.5])
  
  return(total_electoral_votes)
})

simulation_df <- data.frame(total_electoral_votes = simulation_results)

summary(simulation_df$total_electoral_votes)
below_270 <- sum(simulation_df$total_electoral_votes < 270)
below_270/n_simulations

ggplot(simulation_df, aes(x = total_electoral_votes)) +
  geom_histogram(binwidth = 10, fill = "blue", color = "black", alpha = 0.7) +
   geom_vline(xintercept = 270, color = "red", linetype = "dashed", size = 1) +
  labs(
    title = "Distribution of Republican Electoral College Votes",
    x = "Electoral College Votes",
    y = "Frequency"
  ) +
  theme_minimal()
```
This simulation shows Democrats winning the presidency 2.35%. 

# past election data for covariance matrix (excluding diagonal)
  

```{r}
# past data for covariance matrix
past_data <- read.csv("1976-2020-president.csv")
head(past_data)
```


```{r}
past_data <- past_data[, (names(past_data) %in% c("year", "state","candidatevotes"  ,"party_simplified"))]

head(past_data)
```


```{r}
# new dataset with Rep_percent, state, year
library(dplyr)
new_data <- past_data %>%
  filter(party_simplified %in% c("DEMOCRAT", "REPUBLICAN")) %>%
  group_by(year, state) %>%
  summarise(
    democrat_votes = sum(candidatevotes[party_simplified == "DEMOCRAT"]),
    republican_votes = sum(candidatevotes[party_simplified == "REPUBLICAN"]),
    republican_proportion = republican_votes / (democrat_votes + republican_votes),
    .groups = "drop"
  ) %>%
  select(year, state, republican_proportion)

head(new_data)
```



```{r,  fig.width=6, fig.height=6}
ggplot(new_data, aes(x = year, y = republican_proportion)) +
  geom_point(size = 2, alpha = 0.7) +
  facet_wrap(~ state) +
  theme_minimal()
```
Lets choose the past four elections for our covariance matrix

```{r}
new_data <- new_data %>%
  filter(year %in% c(2020, 2016, 2012, 2008, 2004))
```




```{r}
first_51_states <- new_data$state[1:51 ]
first_51_states <- data.frame(
  new_states = new_data$state[1:51],
  sim_states = sim_data$State[1:51]
)

# View the result
print(first_51_states)
```
They match. 

```{r}
# create proportion matrix col1 is Alabama, col2 alaska, etc.
library(tidyr)

wide_data <- new_data %>%
  pivot_wider(
    names_from = state,            
    values_from = republican_proportion  
  )

head(wide_data)
```

```{r}
library(dplyr)
numerical_data <- wide_data %>%
  dplyr::select(-year)

cov_matrix <- cov(numerical_data, use = "pairwise.complete.obs")

#print(cov_matrix)
```




```{r}
# monte carlo simulation
library(MASS)  

set.seed(1)  
n_simulations <- 10000

simulation_results <- replicate(n_simulations, {
  # multivariate normal distribution
  simulated_proportions <- mvrnorm(1, mu = sim_data$simulated_proportion, Sigma = cov_matrix)
  
  simulated_proportions <- pmin(pmax(simulated_proportions, 0), 1)
  
  total_electoral_votes <- sum(sim_data$Elec_votes[simulated_proportions > 0.5])
  
  return(total_electoral_votes)
})

simulation_df <- data.frame(total_electoral_votes = simulation_results)

summary(simulation_df$total_electoral_votes)

below_270 <- sum(simulation_df$total_electoral_votes < 270)
below_270/n_simulations

library(ggplot2)
ggplot(simulation_df, aes(x = total_electoral_votes)) +
  geom_histogram(binwidth = 10, fill = "blue", color = "black", alpha = 0.7) +
   geom_vline(xintercept = 270, color = "red", linetype = "dashed") +
  labs(
    title = "Distribution of Republican Electoral College Votes",
    x = "Electoral College Votes",
    y = "Frequency"
  ) +
  theme_minimal()

```
This simulation show Democrats winning 21.83%


