---
title: '2'
author: "Grant Corkren"
date: "2023-04-28"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
setwd("~/MY_R")
library(openxlsx)
library(TidyDensity)
library(aTSA)
library(itsmr)
library(forecast)
library(uroot)
library(seastests)

A= read.xlsx("lag.xlsx")

A[1,6]=0
A[1,7]=0
A[2,7]=0
A[1,8]=0
A[2,8]=0
A[3,8]=0
level <- A[,c(2,4)]
ind <- A[,2]

level <- level[!is.na(ind),]


head(test)
```


```{r}
tsData <- ts(level[,1], start = c(2011,1), frequency = 365)
tstrain<-window(tsData, start=c(2011,1), end= c(2019,364))
tstest<-window(tsData, start=c(2020,1))

```
```{r}
precip<-ts(A[,5], start = c(2011,1), frequency = 365)
ts.plot(precip)
```


```{r}
naive = snaive(tstrain,  h=length(tstest))
accuracy(naive, tstest)
```
```{r}
plot(tsData, col="blue", xlab="Year", ylab="waterlevel", main="Seasonal naive forecast",type='l')
lines(naive$mean, col="red", lwd=2)
```
```{r}
ets_model=ets(tstrain, allow.multiplicative.trend = TRUE)
summary(ets_model)
```

```{r}
ets_forecast=forecast(ets_model, h=length(tstest))
accuracy(ets_forecast, tstest)
plot(tsData, col="blue", xlab="Year", ylab="waterlevel", main="exponential forecast",type='l')
lines(ets_forecast$mean, col="red", lwd=2)
```


```{r}
tbats_model = tbats(tstrain)
tbats_forecast = forecast(tbats_model, h=length(tstest))
accuracy(tbats_forecast, tstest)
plot(tsData, col="blue", xlab="Year", ylab="waterlevel", main="tbats forecast",type='l')
lines(tbats_forecast$mean, col="red", lwd=2)
```
```{r}
tbats_model
```


```{r}
b<-as.matrix(A[1:3284,6:8])
arima_optimal<-auto.arima(tstrain, xreg = b, max.p = 5, max.q = 5,
                          max.P = 365, d = 1, max.Q = 365, D = 1, stationary = TRUE, seasonal = TRUE)
arima_optimal
#arima_optforecast= forecast(arima_optimal, h=length(tstest))
```








```{r}
arima_opt<-Arima(tstrain,order = c(5,0,3) ,xreg = b )
arima_optforecast= forecast(arima_opt, h=1118, xreg = c)
```

```{r}
accuracy(arima_optforecast, tstest)
plot(tsData, col="blue", xlab="Year", ylab="waterlevel", main="arima optimal forecast",type='l')
lines(arima_optforecast$mean, col="red", lwd=2)
```



```{r}
library(tseries)

library(parallel)
library(MSwM)
b<-as.matrix(b)
#Model with only intercept
mod<-lm(tstrain ~ b)
summary(mod)
#Fit regime-switching model
msar1=msmFit(mod, k=2, sw=c(T,T,T,T,T,T,T), p=2)
msar2= msmFit(mod, k=2, sw=c(T,T,T,T,T,T,T,T), p=3)
summary(msar1)
summary(msar2)

```
```{r}
mod<-lm(tstrain ~ 1)
msar2= msmFit(mod, k=2, sw=c(T,T,T,T,T), p=3)
summary(msar2)
```









plot(mod.mswm)
```{r}
plotDiag(msar1)
plotDiag(msar2)
```


library(mhsmm)
J<-2
initial <- rep(1/J,J)


init0 <- rep(1/J,J)
P0 <- matrix(1/J,nrow=J,ncol=J)
b0 <- list(mu=c(0,0),sigma=c(1,1))
startval <- hmmspec(init=init0, trans=P0,parms.emission=b0,dens.emission=dnorm.hsmm) 
h1 = hmmfit(tstrain,startval,mstep=mstep.norm)

plot(h1$loglik,type='b',ylab='Log-likelihood',xlab='Iteration')
summary(h1)



mean(test$s!=predict(h1,tstest)$s)
predhmm=predict(h1,tstest)
accuracy(predhmm, tstest)



```{r}
c<-as.matrix(A[1:1118,6:8])
newarima<-stlm(tstrain, s.window = "periodic", method =  "arima",  biasadj = FALSE, xreg = b)

predicted<-forecast(newarima, h=length(tstest), newxreg=c)
accuracy(predicted, tstest)

```
```{r}
newarima1$model
```



```{r}
plot(tsData, col="blue", xlab="Year", ylab="waterlevel", main="",type='l')
lines(predicted$mean, col="red", lwd=2)
```


```{r}
newarima1<-stlm(tstrain, s.window = 365, method =  "arima",lambda = TRUE,  biasadj = TRUE)

predicted<-forecast(newarima1, h=length(tstest))
accuracy(predicted, tstest)

```
```{r}
plot(tsData, col="blue", xlab="Year", ylab="waterlevel", main="",type='l')
lines(predicted$mean, col="red", lwd=2)

```
```{r}

```



```{r}
newets<-stlm(tstrain, s.window = "periodic", method =  "ets",
  etsmodel = "ZZN", lambda = "auto", biasadj = FALSE)

predicted<-forecast(newets, h=length(tstest) )
accuracy(predicted, tstest)

```

```{r}
newets$model
```




```{r}
plot(tsData, col="blue", xlab="Year", ylab="waterlevel", main="Exponential seasonally adjusted forecast",type='l')
lines(predicted$mean, col="red", lwd=2)

```


Forecasts of STL objects are obtained by applying a non-seasonal forecasting method to the seasonally adjusted data and re-seasonalizing using the last year of the seasonal component.



