---
title: 'Groundwater Level Analysis'
author: "Grant Corkren"
date: "2023-04-28"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(warn = -1)
```


```{r}
setwd("~/MY_R")
library(openxlsx)
library(TidyDensity)
library(aTSA)
library(itsmr)
library(forecast)

library(readxl)
data <- read_excel("uagw1.xlsx")


head(data)
ts_water <- ts(data$waterlevel, frequency = 365)

train_data <- ts_water[1:3000]
test_data <- ts_water[3001:4403]

```
```{r}
tail(data)
```





```{r}
library(ggplot2)
ggplot(data, aes(x = DATE)) + 
  geom_line(aes(y = waterlevel, color = "Water Level")) + 
  labs(y = "Values", title = "Daily Water Level")
ggplot(data, aes(x = DATE)) + 
  geom_line(aes(y = precip, color = "Rain")) +
  labs(y = "Values", title = "Daily Rain")

```

```{r}
library(zoo)
data$WaterLevel_MA <- rollmean(data$waterlevel, k = 30, fill = NA)
data$Rain_MA <- rollmean(data$precip, k = 30, fill = NA)

ggplot(data, aes(x = DATE)) + 
  geom_line(aes(y = WaterLevel_MA, color = "Water Level MA")) + 
  labs(y = "Moving Average", title = "30-Day Moving Average of Water Level")
ggplot(data, aes(x = DATE)) + 
  geom_line(aes(y = Rain_MA, color = "Rain MA")) +
  labs(y = "Moving Average", title = "30-Day Moving Average of Rain")

```

```{r}

# Use ses with alpha = smoothing parameter
data$WaterLevel_EWMA <- ses(data$waterlevel, alpha = 0.3)$fitted
data$Rain_EWMA <- ses(data$precip, alpha = 0.3)$fitted

ggplot(data, aes(x = DATE)) + 
  geom_line(aes(y = WaterLevel_EWMA, color = "Water Level EWMA")) + 
  labs(y = "Exponentially Weighted Moving Average", title = "Exponentially Weighted Moving Average of Water Level")
ggplot(data, aes(x = DATE)) + 
  geom_line(aes(y = Rain_EWMA, color = "Rain EWMA")) +
  labs(y = "Exponentially Weighted Moving Average", title = "Exponentially Weighted Moving Average of Rain")
```



```{r}
decomposition <- stl(ts_water, s.window = "periodic")
plot(decomposition)
```
Decomposition shows no clear trend, but shows a clear sign of seasonality. The first method chosen will choose the seasonality component and subtract it from the the time series data. The seasonality will be added back after the models have been fitted and forcasted. 



```{r}
data$water_adj <- data$waterlevel - decomposition$time.series[, "seasonal"] 
ts_season <- ts(decomposition$time.series[, "seasonal"])
ts_adj <- ts(data$waterlevel) - ts_season
ts_train_adj <- ts_adj[1:3000]
test_test_adj <- ts_adj[3001:4403]

ggplot(data, aes(x = DATE)) + 
  geom_line(aes(y = water_adj, color = "Water Level Adjusted")) + 
  labs(y = "Values", title = "Daily Water Level Adjusted")
```

```{r}
library(forecast)
library(tseries)
naive_model <- naive( train_data, h = length(test_data))


naive_forecast = forecast(naive_model, h=length(test_data))
naive_forecast$mean = naive_forecast$mean + ts_season[3001:4403]
accuracy(naive_forecast, test_data)
plot(train_data, col="blue", xlab="Year", ylab="waterlevel", main="naive forecast",type='l')
lines(naive_forecast$mean, col="red", lwd=2)


```


ARIMA or ARMA will be used throguh auto.arima function.
Is it stationary 

```{r}
library(tseries)
adf_test <- adf.test(data$water_adj)
print(adf_test)
```
It is stationary. 


```{r}
acf(data$water_adj, main = "ACF of Water Level")
pacf(data$water_adj, main = "PACF of Water Level")
```
```{r}
library(forecast)
xreg_train = data$precip[1:3000]
xreg_test = data$precip[3001:4403]

# ARIMA model with Rain as an external regressor

arima_with_rain <- auto.arima(ts_train_adj, xreg = xreg_train)
summary(arima_with_rain)
arima_forecast = forecast(arima_with_rain, h=length(test_data), xreg= xreg_test)
arima_forecast$mean = arima_forecast$mean + ts_season[3001:4403]
  
  
accuracy(arima_forecast, test_data)
plot(train_data, col="blue", xlab="Year", ylab="waterlevel", main="ARIMA with Rain forecast",type='l')
lines(arima_forecast$mean, col="red", lwd=2)
```
Are baseline model is the naive model with seasonal adjustment which has a mase test 9.7. This arima model has a better mase. 



```{r}
checkresiduals(arima_with_rain)
```
p-value is good but variance is not constant. 



rain moving average as regressor
```{r}
# ARIMA model with EWMA of Rain as an external regressor
arima_with_ewma_rain <- auto.arima(ts_train_adj, xreg = xreg_train)
summary(arima_with_ewma_rain)


arima_forecast = forecast(arima_with_ewma_rain, h=length(test_data), xreg= xreg_test)
arima_forecast$mean = arima_forecast$mean + ts_season[3001:4403]
  
  
accuracy(arima_forecast, test_data)
plot(train_data, col="blue", xlab="Year", ylab="waterlevel", main="ARIMA with Rain forecast",type='l')
lines(arima_forecast$mean, col="red", lwd=2)
```
 

The best arima model is rain with arima(3,1,2) but has problems repeating periods of increased variance. 




Method #2
The Analysis and models show that are is a clear change in variation previous tried. The increase of variation occurs in a period manner which may be explain by bringing in new variables that correlate with the time period. The variables that could explain this are temperature, windspeed, waterlevel of  lake Tucsaloosa. 

 
```{r}
library(dplyr)
data2 <- read.csv('3874603.csv', header=TRUE)
data2 <- data2 %>% filter(STATION == "USW00093806")
data2 <- data2[, !(names(data2) %in% c("NAME", "STATION","TAVG"))]
head(data2)
```

Read in lake waterlevel 
```{r}
head(data)
```
Joining the datasets 
```{r}
data$DATE <- as.Date(data$DATE)
data2$DATE <- as.Date(data2$DATE)
data3 <- merge(data,data2 , by = "DATE")
head(data3)
tail(data3)
```

```{r}
nrow(data3)
```

```{r}
ts_water <- ts(data3$waterlevel, frequency = 365)

train_data <- ts_water[1:3000]
test_data <- ts_water[3001:4343]
```

```{r}

decomposition <- stl(ts_water, s.window = "periodic")
plot(decomposition)
```

```{r}
data3$water_adj <- data3$waterlevel - decomposition$time.series[, "seasonal"] 
ts_season <- ts(decomposition$time.series[, "seasonal"])
ts_adj <- ts(data3$waterlevel) - ts_season
ts_train_adj <- ts_adj[1:3000]
test_test_adj <- ts_adj[3001:4343]

ggplot(data3, aes(x = DATE)) + 
  geom_line(aes(y = water_adj, color = "Water Level Adjusted")) + 
  labs(y = "Values", title = "Daily Water Level Adjusted")

ts_noadj <- ts(data3$waterlevel)
ts_train_noadj <- ts_adj[1:3000]
test_test_noadj <- ts_adj[3001:4343]
```

```{r}
ggplot(data3, aes(x = DATE)) + 
  geom_line(aes(y = AWND, color = "Average Wind Speed")) + 
  labs(y = "Values", title = "Average Wind Speend")
ggplot(data3, aes(x = DATE)) + 
  geom_line(aes(y = TMAX, color = "Temp Max")) +
  labs(y = "Values", title = "Temperature Maximum")
ggplot(data3, aes(x = DATE)) + 
  geom_line(aes(y = TMIN, color = "Temp Min")) +
  labs(y = "Values", title = "Temperature Minimum")
```


```{r}
sum(is.na(data3$AWND))  
sum(is.na(data3$TMAX))
sum(is.na(data3$TMIN))
sum(is.na(data3$TDIFF))
```

```{r}
data3$AWND <- na.approx(data3$AWND, na.rm = FALSE)
data3$TMAX <- na.approx(data3$TMAX, na.rm = FALSE)
data3$TMIN <- na.approx(data3$TMIN, na.rm = FALSE)
```

Differencing of max and min temp
```{r}
data3$TDIFF <- data3$TMAX - data3$TMIN
ggplot(data3, aes(x = DATE)) + 
  geom_line(aes(y = TDIFF, color = "Temp DIFF")) +
  labs(y = "Values", title = "Temperature Difference")
```


```{r}
data3$AWND_EWMA <- ses(data3$AWND, alpha = 0.3)$fitted
data3$TMAX_EWMA <- ses(data3$TMAX, alpha = 0.3)$fitted
data3$TMIN_EWMA <- ses(data3$TMIN, alpha = 0.3)$fitted
data3$TDIFF_EWMA <- ses(data3$TDIFF, alpha = 0.3)$fitted
ggplot(data3, aes(x = DATE)) + 
  geom_line(aes(y = AWND_EWMA, color = "Wind EWMA")) +
  labs(y = "Exponentially Weighted Moving Average", title = "Exponentially Weighted Moving Average of Average Wind Speed")
ggplot(data3, aes(x = DATE)) + 
  geom_line(aes(y = TMAX_EWMA, color = "Temp Max EWMA")) +
  labs(y = "Exponentially Weighted Moving Average", title = "Exponentially Weighted Moving Average of Temperature Maximum")
ggplot(data3, aes(x = DATE)) + 
  geom_line(aes(y = TMIN_EWMA, color = "Temp Min EWMA")) +
  labs(y = "Exponentially Weighted Moving Average", title = "Exponentially Weighted Moving Average of Temperature Minimum")
ggplot(data3, aes(x = DATE)) + 
  geom_line(aes(y = TDIFF_EWMA, color = "Temp DIFF EWMA")) +
  labs(y = "Exponentially Weighted Moving Average", title = "Exponentially Weighted Moving Average of Temperature Difference")
```

Auto Arma not engineered regressors
```{r}
library(forecast)
# Define the target time series
Precipreg_train = data3$precip[1:3000]
Precipreg_test = data3$precip[3001:4343]
AWNDreg_train = data3$AWND[1:3000]
AWNDreg_test = data3$AWND[3001:4343]
TMAXreg_train = data3$TMAX[1:3000]
TMAXreg_test = data3$TMAX[3001:4343]
TMINreg_train = data3$TMIN[1:3000]
TMINreg_test = data3$TMIN[3001:4343]
#TDIFFreg_train = data3$TDIFF[1:3000]
#TDIFFreg_test = data3$TDIFF[3001:4343]

xreg_train<- cbind(Precipreg_train, AWNDreg_train, TMAXreg_train, TMINreg_train)
xreg_test<- cbind(Precipreg_test, AWNDreg_test, TMAXreg_test, TMINreg_test)
# ARIMA model with Rain as an external regressor

arima_with_regs <- auto.arima(ts_train_adj, xreg = xreg_train)
summary(arima_with_regs)
arima_forecast = forecast(arima_with_regs, h=length(test_data), xreg= xreg_test)
arima_forecast$mean = arima_forecast$mean + ts_season[3001:4343]
  
  
accuracy(arima_forecast, test_data)
plot(train_data, col="blue", xlab="Year", ylab="waterlevel", main="ARIMA with Multiple Regessors forecast",type='l')
lines(arima_forecast$mean, col="red", lwd=2)
```

```{r}
checkresiduals(arima_with_regs)
```
Seasonality with a period of increase variation still the analysis. Temperature and Wind have little effect on the waterlevel. 


```{r}
library(forecast)
# Define the target time series
Precipreg_train = data3$precip[1:3000]
Precipreg_test = data3$precip[3001:4343]
AWNDreg_train = data3$AWND_EWMA[1:3000]
AWNDreg_test = data3$AWND_EWMA[3001:4343]
TMAXreg_train = data3$TMAX_EWMA[1:3000]
TMAXreg_test = data3$TMAX_EWMA[3001:4343]
TMINreg_train = data3$TMIN_EWMA[1:3000]
TMINreg_test = data3$TMIN_EWMA[3001:4343]
TDIFFreg_train = data3$TDIFF_EWMA[1:3000]
TDIFFreg_test = data3$TDIFF_EWMA[3001:4343]

xreg_train<- cbind(Precipreg_train, AWNDreg_train, TMAXreg_train, TMINreg_train, TDIFFreg_train)
xreg_test<- cbind(Precipreg_test, AWNDreg_test, TMAXreg_test, TMINreg_test, TDIFFreg_test)
# ARIMA model with Rain as an external regressor

arima_with_regs <- auto.arima(ts_train_adj, xreg = xreg_train)
summary(arima_with_regs)
arima_forecast = forecast(arima_with_regs, h=length(test_data), xreg= xreg_test)
arima_forecast$mean = arima_forecast$mean + ts_season[3001:4343]
  
  
accuracy(arima_forecast, test_data)
plot(train_data, col="blue", xlab="Year", ylab="waterlevel", main="ARIMA with Multiple Regessors forecast",type='l')
lines(arima_forecast$mean, col="red", lwd=2)
```

```{r}
checkresiduals(arima_with_regs)
```

Best MASE but still has the same problem with an unexplained increase in variance. 

Trying with no seasonality adjustment. 
```{r}
library(forecast)
# Define the target time series
Precipreg_train = data3$precip[1:3000]
Precipreg_test = data3$precip[3001:4343]
AWNDreg_train = data3$AWND_EWMA[1:3000]
AWNDreg_test = data3$AWND_EWMA[3001:4343]
TMAXreg_train = data3$TMAX_EWMA[1:3000]
TMAXreg_test = data3$TMAX_EWMA[3001:4343]
TMINreg_train = data3$TMIN_EWMA[1:3000]
TMINreg_test = data3$TMIN_EWMA[3001:4343]

xreg_train<- cbind(Precipreg_train, AWNDreg_train, TMAXreg_train, TMINreg_train)
xreg_test<- cbind(Precipreg_test, AWNDreg_test, TMAXreg_test, TMINreg_test)
# ARIMA model with Rain as an external regressor

arima_with_regs <- auto.arima(ts_train_noadj, xreg = xreg_train)
summary(arima_with_regs)
arima_forecast = forecast(arima_with_regs, h=length(test_data), xreg= xreg_test)
  
  
accuracy(arima_forecast, test_data)
plot(train_data, col="blue", xlab="Year", ylab="waterlevel", main="ARIMA with Multiple Regessors forecast",type='l')
lines(arima_forecast$mean, col="red", lwd=2)
```





