---
title: "Groundwater Level Analysis"
author: "Grant Corkren"
date: "2023-04-28"
output:
  pdf_document: default
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(warn = -1)
```

This project was done in a consulting class where I consulted the Geological Survey of Alabama on the height of water in a well located in Tuscaloosa County. I was tasked with finding if their was a trend and seasonality. Additionally I was asked if it could be forecasted and if rain effected well water height. 

This analysis includes time series decomposition to answer the question of trend and seasonality. Their is no trend but their is seasonality. For modeling ARIMA models their needs to be stationarity so a seasonal difference was taken. It also includes an ARMA model with rain as a variable to answer whether or not rain effects well water level. The resulting coefficient of rain in that ARIMA model was 0.02 and was statistically significant. 

ARIMA models were applied to answer the question of forecasting. Rain is not used as a variable as that would need to be forecasted also or used conditionally.

Ultimately all of the forecasting models suffered from a non constant variance. A second analysis has been done to explore the time series data further and apply GARCH models. 


```{r, include=FALSE}
setwd("~/MY_R")
library(openxlsx)
library(TidyDensity)
library(aTSA)
library(itsmr)
library(forecast)

library(readxl)
data <- read_excel("uagw1.xlsx")


head(data)
ts_water <- ts(data$waterlevel, frequency = 365)

train_data <- ts_water[1:3000]
test_data <- ts_water[3001:4403]

```
```{r}
tail(data)
```


```{r}
library(ggplot2)
# Analyze rain
ggplot(data, aes(x = precip)) +
  geom_histogram(binwidth = .25, fill = "blue", color = "white") +
  labs(title = "Histogram of Rain", x = "Value", y = "Frequency") +
  theme_minimal()

summary(data$precip)

```



```{r}
library(ggplot2)
ggplot(data, aes(x = DATE)) + 
  geom_line(aes(y = waterlevel, color = "Water Level")) + 
  labs(y = "Values", title = "Daily Water Level")
ggplot(data, aes(x = DATE)) + 
  geom_line(aes(y = precip, color = "Rain")) +
  labs(y = "Values", title = "Daily Rain")

```



```{r, include=FALSE}

# Use ses with alpha = smoothing parameter
data$WaterLevel_EWMA <- ses(data$waterlevel, alpha = 0.3)$fitted
data$Rain_EWMA <- ses(data$precip, alpha = 0.3)$fitted

ggplot(data, aes(x = DATE)) + 
  geom_line(aes(y = WaterLevel_EWMA, color = "Water Level EWMA")) + 
  labs(y = "Exponentially Weighted Moving Average", title = "Exponentially Weighted Moving Average of Water Level")
ggplot(data, aes(x = DATE)) + 
  geom_line(aes(y = Rain_EWMA, color = "Rain EWMA")) +
  labs(y = "Exponentially Weighted Moving Average", title = "Exponentially Weighted Moving Average of Rain")
```



```{r}
decomposition <- stl(ts_water, s.window = "periodic")
plot(decomposition)
```
Decomposition shows no clear trend, but shows a clear sign of seasonality. The first method chosen will choose the seasonality component and subtract it from the the time series data. The seasonality will be added back after the models have been fitted and forcasted. 



```{r}
data$water_adj <- data$waterlevel - decomposition$time.series[, "seasonal"] 
ts_season <- ts(decomposition$time.series[, "seasonal"])
ts_adj <- ts(data$waterlevel) - ts_season
ts_train_adj <- ts_adj[1:3000]
test_test_adj <- ts_adj[3001:4403]

ggplot(data, aes(x = DATE)) + 
  geom_line(aes(y = water_adj, color = "Water Level Adjusted")) + 
  labs(y = "Values", title = "Daily Water Level Adjusted")
```

```{r}
# add actual Too
library(forecast)
library(tseries)
naive_model <- naive( train_data, h = length(test_data))


naive_forecast = forecast(naive_model, h=length(test_data))
naive_forecast$mean = naive_forecast$mean + ts_season[3001:4403]
accuracy(naive_forecast, test_data)
plot(train_data, col="blue", xlab="Year", ylab="waterlevel", main="naive forecast",type='l')
lines(naive_forecast$mean, col="red", lwd=2)


```


ARIMA or ARMA will be used throguh auto.arima function.
Is it stationary 

```{r}
library(tseries)
adf_test <- adf.test(data$water_adj)
print(adf_test)
```
Using the Augmented Dickey-Fuller Test shows the seasonal differenced time series is stationary.

```{r}
acf(data$water_adj, main = "ACF of Water Level")
pacf(data$water_adj, main = "PACF of Water Level")
```
## Rain as exogenous regressor
This model is not intended for forecasting although it could be with conditional forecasting. This model is intended to answer whether or not rain has an effect on the waterlevel of the well. 
```{r}
library(forecast)
xreg_train = data$precip[1:3000]
xreg_test = data$precip[3001:4403]

# ARIMA model with Rain as an external regressor

arima_with_rain <- auto.arima(ts_train_adj, xreg = xreg_train)


```
Are baseline model is the naive model with seasonal adjustment which has a mase test 9.7. This arima model has a better mase. 

```{r}
# computing p values
model <- arima(ts_train_adj, order = c(3, 1, 3), xreg = xreg_train)

coefs <- model$coef  
se <- sqrt(diag(model$var.coef))


t_values <- coefs / se

# Computing p-values (two-tailed test)
p_values <- 2 * (1 - pt(abs(t_values), df = length(ts_train_adj) - length(coefs)))


results <- data.frame(Coefficient = coefs, Std_Error = se, t_value = t_values, p_value = p_values)
print(results)

```
All p-values are statistically signficant. 

```{r}
checkresiduals(arima_with_rain)
```
For a first model it performs fairly well but suffers from a non-constant variance in residuals. My best guess is that the winter period, which is more volatile, is contributing to that. 


## Forecasting models

# ARIMA

```{r}
#ARMA without rain for forecasting
arima_no_rain <- auto.arima(ts_train_adj)
summary(arima_no_rain)
arima_forecast = forecast(arima_no_rain, h=length(test_data))
arima_forecast$mean = arima_forecast$mean + ts_season[3001:4403]
  
  
accuracy(arima_forecast, test_data)
plot(train_data, col="blue", xlab="Year", ylab="waterlevel", main="ARIMA with Rain forecast",type='l')
lines(arima_forecast$mean, col="red", lwd=2)


```
```{r}
# computing p values
model <- arima(ts_train_adj, order = c(2, 1, 2))

coefs <- model$coef  
se <- sqrt(diag(model$var.coef))


t_values <- coefs / se

# Computing p-values (two-tailed test)
p_values <- 2 * (1 - pt(abs(t_values), df = length(ts_train_adj) - length(coefs)))


results <- data.frame(Coefficient = coefs, Std_Error = se, t_value = t_values, p_value = p_values)
print(results)
```
The AR3 was not statistically significant and was removed.

```{r}
summary(model)
arima_forecast = forecast(model, h=length(test_data))
arima_forecast$mean = arima_forecast$mean + ts_season[3001:4403]
  
  
accuracy(arima_forecast, test_data)
plot(train_data, col="blue", xlab="Year", ylab="waterlevel", main="ARIMA with Rain forecast",type='l')
lines(arima_forecast$mean, col="red", lwd=2)
```

```{r}
checkresiduals(model)
```

Given the seasonal naive model has a testing MASE of 9.7, it can be concluded that seasonally adjust ARIMA model with the order (Auto Regressive order of 2, Integration order of 1, Moving Average order of 2) is a better model. This model does suffer from a non constant variance of residuals. Examining the residuals of the model shows that the residuals are 


## Conclusion

The analysis shows that their is seasonality but no trend. It also shows that rain does effect well water height, increasing the well water height by 0.02 inches for every inch of rain. The ARIMA model created is suitable for forecasting but suffers non constant variance of residuals, periods of higher inaccuracy. ARCH and GARCH may need to be applied to address this. 





